---
title: "Simulating layered Brownian bridges"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{layered_brownian_bridges}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(layeredBB)
```

## Simulating layer information

In many settings, we want to simulate Brownian motion together with *layer information*. Layer information determines a compact spatial interval which constrains the path over a given time interval. There are several different layer information available, but in this package, we only implement the *Bessel layer* approach outlined in (Beskos, Papaspiliopoulos, and Roberts 2008) and in Algorithm 14 of (Pollock, Johansen, and Roberts 2016).

The general idea of constructing layer information for Brownian bridge sample paths is that finite dimensional subsets of Brownian bridge sample paths can be simulated jointly with information regarding the interval in which its constrained by partitioning the path space with an arbitrary increasing sequence $\{a_{i}\}_{i \geq 0}, a_{0} = 0$, which radiates outwards from the interval $[(x \land y), (x \lor y)]$. We term this particular layer construction, the \emph{Bessel layer} - the $i$th Bessel layer is defined as
$$
\mathcal{I}_{i} = [(x \land y) - a_{i}, (x \lor y) + a_{i}].
$$

To simulate a Bessel layer, we use the `bessel_layer_simulation` function, where we must pass in the usual start and end points of the Brownian bridge with `x`, `y`, `s` and `t` denoting the start point, end point, start time, end time of the bridge respectively. The size of layers is given by $\sqrt{t-s}*mult$, where $mult$ is the multiplier passed into the `mult` argument (which is $1$ by default). For instance, we can simulate a Bessel layer between $x=-0.4$ to $y=0.2$ in time $[0,1]$ using the following code:

```{r, fig.align='center', fig.dim=c(6,4)}
set.seed(2021)
x <- -0.4
y <- 0.2
s <- 0
t <- 1
bes_layer <- bessel_layer_simulation(x = x, y = y, s = s, t = t, mult = 0.2)
# plot simulated Bessel layer information
plot(x = c(s,t), y = c(x,y), ylim = c(-1, 1), pch = 20, lwd = 4,
     xlab = '', ylab = '', xaxt = 'n', yaxt = 'n')
mtext('Time', 1, 2.75, font = 2, cex = 1.5)
mtext('W', 2, 2.75, font = 2, cex = 1.5)
axis(1, at=seq(0, 1, 0.2), labels=c("0.0", seq(0.2, 0.8, 0.2), "1.0"), font = 2, cex = 1.5)
axis(1, at=seq(0, 1, 0.1), labels=rep("", 11), lwd.ticks = 0.5)
axis(2, at=seq(-1, 1, 0.5), labels=seq(-1, 1, 0.5), font = 2, cex = 1.5)
axis(2, at=seq(-1, 1, 0.1), labels=rep("", 21), lwd.ticks = 0.5)
abline(h=c(bes_layer$L, bes_layer$U), lwd = 3)
abline(h=c(bes_layer$l, bes_layer$u), lwd = 3, lty = 2)
```

In this above plot, the solid lines denote the interval in space where the Brownian bridge is definitely constrained, so the sample path falls entirely between these lower and upper lines at $L=-1$ and $U=0.8$. The dashed lines indicate the interval which does not constrain the path, as at some point, the path can will outside these dotted lines. Having a look at what this function returns:
```{r}
print(bes_layer)
```

Here `bes_layer$L` and `bes_layer$U` returns the 'hard' bounds for the Bessel layer (i.e. the path is completely constrained between these two points) whereas `bes_layer$l` and `bes_layer$u` return the 'soft' bounds for the Bessel layer (i.e. the path can fall outside these bounds).

We can then simulate a Brownian bridge conditional on the Bessel layer information using the `layered_brownian_bridge` function by passing in a simulated Bessel layer into the `bessel_layer` argument and the times to which we want to simulate in the `times` argument.

## Simulating layered Brownian bridges

```{r, fig.align='center', fig.dim=c(6,4)}
layered_bb <- layered_brownian_bridge(x = x,
                                      y = y,
                                      s = s,
                                      t = t,
                                      bessel_layer = bes_layer,
                                      times = seq(s, t, 0.01))
# plot simulated layered Brownian bridge
plot(x = c(s,t), y = c(x,y), ylim = c(-1, 1), pch = 20, lwd = 4,
     xlab = '', ylab = '', xaxt = 'n', yaxt = 'n')
lines(x = layered_bb$full_path['time',], y = layered_bb$full_path['X',], lwd = 3)
mtext('Time', 1, 2.75, font = 2, cex = 1.5)
mtext('W', 2, 2.75, font = 2, cex = 1.5)
axis(1, at=seq(0, 1, 0.2), labels=c("0.0", seq(0.2, 0.8, 0.2), "1.0"), font = 2, cex = 1.5)
axis(1, at=seq(0, 1, 0.1), labels=rep("", 11), lwd.ticks = 0.5)
axis(2, at=seq(-1, 1, 0.5), labels=seq(-1, 1, 0.5), font = 2, cex = 1.5)
axis(2, at=seq(-1, 1, 0.1), labels=rep("", 21), lwd.ticks = 0.5)
abline(h=c(bes_layer$L, bes_layer$U), lwd = 3)
abline(h=c(bes_layer$l, bes_layer$u), lwd = 3, lty = 2)
```
We can see that our sample path is indeed constrained between the two solid lines but has exceeded the dashed lines.

## Multidimensional layered Brownian bridges

The `multi_bessel_layer_simulation` and `multi_layered_brownian_bridge` functions implement the above functions recursively for each dimension of the space. The difference here is that we must pass in vectors for the arguments `x` and `y`. To simulate the Bessel layer information for a two-dimensional Brownian bridge from $x=(0,0)$ to $y=(1,1)$ in time $[0,1]$ we can use the following code:

```{r}
set.seed(2021)
x <- c(0,0)
y <- c(1,1)
s <- 0
t <- 1
multi_bes_layer <- multi_bessel_layer_simulation(dim = 2,
                                                 x = x,
                                                 y = y,
                                                 s = s,
                                                 t = t,
                                                 mult = 0.2)
print(multi_bes_layer)
```

We can now see that this function now returns a list of length 2, where each item in the list is the Bessel layer for the corresponding dimension.

```{r, fig.align='center', fig.dim=c(6,10)}
par(mfrow=c(2,1))
# dimension 1
plot(x = c(s,t), y = c(x[1],y[1]), ylim = c(-1, 2), pch = 20, lwd = 4,
     xlab = '', ylab = '', xaxt = 'n', yaxt = 'n', main = 'Dimension 1')
mtext('Time', 1, 2.75, font = 2, cex = 1.5)
mtext('W', 2, 2.75, font = 2, cex = 1.5)
axis(1, at=seq(0, 1, 0.2), labels=c("0.0", seq(0.2, 0.8, 0.2), "1.0"), font = 2, cex = 1.5)
axis(1, at=seq(0, 1, 0.1), labels=rep("", 11), lwd.ticks = 0.5)
axis(2, at=seq(-1, 2, 0.5), labels=seq(-1, 2, 0.5), font = 2, cex = 1.5)
axis(2, at=seq(-1, 2, 0.1), labels=rep("", 31), lwd.ticks = 0.5)
abline(h=c(multi_bes_layer[[1]]$L, multi_bes_layer[[1]]$U), lwd = 3)
abline(h=c(multi_bes_layer[[1]]$l, multi_bes_layer[[1]]$u), lwd = 3, lty = 2)
# dimension 2
plot(x = c(s,t), y = c(x[2],y[2]), ylim = c(-1, 2), pch = 20, lwd = 4,
     xlab = '', ylab = '', xaxt = 'n', yaxt = 'n', main = 'Dimension 2')
mtext('Time', 1, 2.75, font = 2, cex = 1.5)
mtext('W', 2, 2.75, font = 2, cex = 1.5)
axis(1, at=seq(0, 1, 0.2), labels=c("0.0", seq(0.2, 0.8, 0.2), "1.0"), font = 2, cex = 1.5)
axis(1, at=seq(0, 1, 0.1), labels=rep("", 11), lwd.ticks = 0.5)
axis(2, at=seq(-1, 2, 0.5), labels=seq(-1, 2, 0.5), font = 2, cex = 1.5)
axis(2, at=seq(-1, 2, 0.1), labels=rep("", 31), lwd.ticks = 0.5)
abline(h=c(multi_bes_layer[[2]]$L, multi_bes_layer[[2]]$U), lwd = 3)
abline(h=c(multi_bes_layer[[2]]$l, multi_bes_layer[[2]]$u), lwd = 3, lty = 2)
```
Similarly, we can pass in the multidimensional Bessel layer information into the `bessel_layer` argument in the `multi_layered_brownian_bridge` function to simulate a Brownian bridge conditonal on the Bessel layer:

```{r, fig.align='center', fig.dim=c(6,10)}
multi_layered_bb <- multi_layered_brownian_bridge(dim = 2,
                                                  x = x,
                                                  y = y,
                                                  s = s,
                                                  t = t,
                                                  bessel_layers = multi_bes_layer,
                                                  times = seq(s, t, 0.01))
par(mfrow=c(2,1))
# dimension 1
plot(x = c(s,t), y = c(x[1],y[1]), ylim = c(-1, 2), pch = 20, lwd = 4,
     xlab = '', ylab = '', xaxt = 'n', yaxt = 'n', main = 'Dimension 1')
lines(x = multi_layered_bb$full_path[3,], y = multi_layered_bb$full_path[1,], lwd = 3)
mtext('Time', 1, 2.75, font = 2, cex = 1.5)
mtext('W', 2, 2.75, font = 2, cex = 1.5)
axis(1, at=seq(0, 1, 0.2), labels=c("0.0", seq(0.2, 0.8, 0.2), "1.0"), font = 2, cex = 1.5)
axis(1, at=seq(0, 1, 0.1), labels=rep("", 11), lwd.ticks = 0.5)
axis(2, at=seq(-1, 2, 0.5), labels=seq(-1, 2, 0.5), font = 2, cex = 1.5)
axis(2, at=seq(-1, 2, 0.1), labels=rep("", 31), lwd.ticks = 0.5)
abline(h=c(multi_bes_layer[[1]]$L, multi_bes_layer[[1]]$U), lwd = 3)
abline(h=c(multi_bes_layer[[1]]$l, multi_bes_layer[[1]]$u), lwd = 3, lty = 2)
# dimension 2
plot(x = c(s,t), y = c(x[2],y[2]), ylim = c(-1, 2), pch = 20, lwd = 4,
     xlab = '', ylab = '', xaxt = 'n', yaxt = 'n', main = 'Dimension 2')
lines(x = multi_layered_bb$full_path[3,], y = multi_layered_bb$full_path[2,], lwd = 3)
mtext('Time', 1, 2.75, font = 2, cex = 1.5)
mtext('W', 2, 2.75, font = 2, cex = 1.5)
axis(1, at=seq(0, 1, 0.2), labels=c("0.0", seq(0.2, 0.8, 0.2), "1.0"), font = 2, cex = 1.5)
axis(1, at=seq(0, 1, 0.1), labels=rep("", 11), lwd.ticks = 0.5)
axis(2, at=seq(-1, 2, 0.5), labels=seq(-1, 2, 0.5), font = 2, cex = 1.5)
axis(2, at=seq(-1, 2, 0.1), labels=rep("", 31), lwd.ticks = 0.5)
abline(h=c(multi_bes_layer[[2]]$L, multi_bes_layer[[2]]$U), lwd = 3)
abline(h=c(multi_bes_layer[[2]]$l, multi_bes_layer[[2]]$u), lwd = 3, lty = 2)
```
We can just plot this with a two-dimensional plot (where time would be the third axis, but not plotted):

```{r, fig.align='center', fig.dim=c(6,6)}
plot(x = c(x[1], y[1]), y = c(x[2], y[2]), ylim = c(-1, 2), xlim = c(-1, 2), pch = 20, lwd = 4,
     xlab = '', ylab = '', xaxt = 'n', yaxt = 'n')
lines(x = multi_layered_bb$full_path[1,], y = multi_layered_bb$full_path[2,], lwd = 3)
mtext(expression(bold(X[1])), 1, 2.75, font = 2, cex = 1.5)
mtext(expression(bold(X[2])), 2, 2.75, font = 2, cex = 1.5)
axis(1, at=seq(-1, 2, 0.5), labels=seq(-1, 2, 0.5), font = 2, cex = 1.5)
axis(1, at=seq(-1, 2, 0.1), labels=rep("", 31), lwd.ticks = 0.5)
axis(2, at=seq(-1, 2, 0.5), labels=seq(-1, 2, 0.5), font = 2, cex = 1.5)
axis(2, at=seq(-1, 2, 0.1), labels=rep("", 31), lwd.ticks = 0.5)
abline(v=c(multi_bes_layer[[1]]$L, multi_bes_layer[[1]]$U), lwd = 3)
abline(v=c(multi_bes_layer[[1]]$l, multi_bes_layer[[1]]$u), lwd = 3, lty = 2)
abline(h=c(multi_bes_layer[[2]]$L, multi_bes_layer[[2]]$U), lwd = 3)
abline(h=c(multi_bes_layer[[2]]$l, multi_bes_layer[[2]]$u), lwd = 3, lty = 2)
```

### Remark

In the package, there are various functions that are created to implement the algorithm for simulating Bessel layer information and layered Brownian bridges conditional on Bessel layers, for instance:

* delta_coin
* delta_coin_intervals
* eachi
* eadelta
* eadelta_intervals
* eadelta1
* eadelta1_intervals
* eadelta2
* eadelta2_intervals
* eagamma
* eaphi
* eaphi_bar
* eapsi
* easigma
* easigma_bar
* gamma_coin
* gamma_coin_intervals

Each of these functions have documentation to explain how to use them, but unless you are using these to implement other layered Brownian bridge methods, they are not of much interest for you. 

### References

* Beskos, A., O. Papaspiliopoulos, and G. O. Roberts. 2008. “A factorisation of diffusion measure and finite sample path constructions.” Methodology and Computing in Applied Probability 10: 85–104.
* Pollock, M., A. M. Johansen, and G. O. Roberts. 2016. “On the exact and ε-strong simulation of (jump) diffusions.” Bernoulli 22 (2): 794–856.
